Data Pre-Processing/Cleaning Script

### 2023-02-24 code/analysis_code/exploratory_analysis.qmd
part 2 additional work

# Setup

```{r}
#load needed packages. make sure they are installed.
library(here) #for data loading/saving
library(tidyverse)

```

Load the data.

```{r}
#Path to data. Note the use of the here() package and not absolute paths
data_location <- here::here("data","processed_data","processeddata.rds")
#load data
mydata <- readRDS(data_location)
```
#outcome: hospitalization/illness
#predictors:year/month/Food/State/Location
#logistic regression models
#visualizaiton: 
linegraphs for years/month and illness, group by species/IFSAC code
boxplot/violin for top 10 food/state/location with hospitalization and illness
tables for IFSAC/ pathogens

# Data exploration through tables

Showing a bit of code to produce and save a summary table.

This uses a bit of code trickery to get things into the shape of a data frame (for easier saving/showing in manuscript)

```{r}
summary_df = data.frame(do.call(cbind, lapply(mydata, summary)))
print(summary_df)
# another way of looking at the data
dplyr::glimpse(mydata)
# save to file
summarytable_file = here("results", "summarytable.rds")
saveRDS(summary_df, file = summarytable_file)
```

We are saving the results to the `results` folder. Depending on how many tables/figures you have, it might make sense to have separate folders for each. And/or you could have separate folders for exploratory tables/figures and for final tables/figures. Just choose a setup that makes sense for your project and works for you, and provide enough documentation that someone can understand what you are doing.

# Data exploration through figures

```{r}
rawdata1%>%filter(grepl("Confirmed", Status))
food=rawdata1%>%
  filter(!is.na(Food))%>%
  group_by(Food)%>%
  top_n(n=10)
food
species=rawdata1%>%
  filter(!is.na(Species))%>%
  mutate(Species=fct_collapse(Species,Norovirus=c("Norovirus","Norovirus genogroup I","Norovirus genogroup II","Norovirus unknown")))%>%
  mutate(Species=fct_lump(Species,11,other_level = NA))%>%
  count(Species)%>%
  arrange(desc(n))%>%
  filter(!is.na(Species))
species
```

```{r}
p1 <- mydata %>% ggplot(aes(x=Height)) + geom_histogram() 
plot(p1)
figure_file = here("results","height_distribution.png")
ggsave(filename = figure_file, plot=p1) 
```

Now weights.

```{r}
p2 <- mydata %>% ggplot(aes(x=Weight)) + geom_histogram() 
plot(p2)
figure_file = here("results","weight_distribution.png")
ggsave(filename = figure_file, plot=p2) 
```

Now height as function of weight.

```{r}
p3 <- mydata %>% ggplot(aes(x=Height, y=Weight)) + geom_point() + geom_smooth(method='lm')
plot(p3)
figure_file = here("results","height_weight.png")
ggsave(filename = figure_file, plot=p3) 
```

Once more height as function of weight, stratified by sex.

```{r}
p4 <- mydata %>% ggplot(aes(x=Height, y=Weight, color = Sex)) + geom_point() + geom_smooth(method='lm')
plot(p4)
figure_file = here("results","height_weight_stratified.png")
ggsave(filename = figure_file, plot=p4) 
```

# Notes

For your own explorations, tables and figures can be "quick and dirty". As long as you can see what's going on, there is no need to polish them. That's in contrast to figures you'll produce for your final products (paper, report, presentation, website, etc.). Those should look as nice, polished and easy to understand as possible.

>>>>

# Load Libraries

```{r}
# General Data Handling
library(tidyverse)

# For Data Loading and Saving
library(here)
```

# Load Data

## Training Data
```{r}
# Path to Data
data_location <- here(
  "data","processed_data","processed_training_data.rds")
# Import Data
training_data_final <- readRDS(data_location)
```

## Analysis Data
```{r}
# Path to Data
data_location <- here(
  "data","processed_data","processed_analysis_data.rds")
# Import Data
analysis_data_final <- readRDS(data_location)
```

# Data Exploration
## Training Data

The outcome of interest we will explore for the training data is hospitalizations.

```{r}
# Time vs Hospitalizations
ggplot(training_data_final %>% group_by(Year, Pathogen_Type, Hospitalizations) %>% count() %>% mutate(true_hospitalization = n * Hospitalizations), aes(x = Year, y = true_hospitalization, group = Pathogen_Type, color = Pathogen_Type)) + geom_point() + geom_jitter() + labs(x = "Year", y = "Hospitalizations", title = "Scatter Plot of Hospitalizations by Year and Pathogen Type")

## Save Graph
ggsave("time_v_hospitalization_pathogen_type.png", path = "results")

# Pathogen Type + Etiology vs Hospitalization
ggplot(training_data_final %>% group_by(Pathogen_Type, Simplified_Etiology, Hospitalizations) %>% count() %>% mutate(true_hospitalization = n * Hospitalizations), aes(x = Pathogen_Type, y = true_hospitalization, group = Simplified_Etiology, fill = Simplified_Etiology)) + geom_col() + labs(x = "Pathogen Type", y = "Hospitalizations", title = "Stacked Bar Chart of Hospitalizations by Pathogen Type and Etiology")

## Save Graph
ggsave("pathogen_type_etiology_v_hospitalizations.png", path = "results")

## Closer Look at Bacterial Etiologies for Outbreaks
ggplot(training_data_final %>% select(Pathogen_Type, Simplified_Etiology, Hospitalizations) %>% filter(Pathogen_Type == "Bacteria"), aes(x = Simplified_Etiology)) + geom_bar() + labs(x = "Bacteria", y = "Hospitalizations", title = "Bar Chart of Bacteria Frequency")

ggsave("bacterial_etiology_v_hospitalization_frequency.png", path = "results")

# States vs Hospitalization

state_v_hospitalization_training <- training_data_final %>% group_by(State) %>% mutate(all_hospitalizations = sum(Hospitalizations)) %>% count(all_hospitalizations) %>% select(State, all_hospitalizations) %>% arrange(desc(all_hospitalizations))

saveRDS(state_v_hospitalization_training, file = "results/state_v_hospitalization_training.rds")

rmarkdown::paged_table(state_v_hospitalization_training)

# Food Type and Hospitalizations
ggplot(training_data_final %>% filter(`IFSAC Category` != "Multiple") %>% filter(`IFSAC Category` != "Other"), aes(x = `IFSAC Category`, y = Hospitalizations)) + geom_col() + labs(x = "Food Type", y = "Hospitalizations", title = "Hospitalizations by Food Type, Excluding 'Multiple' and 'Other'")

ggsave("food_type_v_hospitalization.png", path = "results")

```
All outbreaks with a high hospitalization count (> 200 cases) are caused by bacteria. It also appears that most outbreaks caused by bacteria are due to Salmonella. California and Minnesota are the two states with the highest total hospitalization count due to foodborne illness from 1998 to 2015.

## Analysis Data
```{r}
# Illnesses Caused by Pathogen Type Over Time
ggplot(analysis_data_final %>% group_by(Year, Pathogen_Type, Illnesses) %>% count() %>% mutate(true_illness_count = n * Illnesses), aes(x = Year, y = true_illness_count, group = Pathogen_Type, color = Pathogen_Type)) + geom_point() + geom_jitter() + labs(x = "Year", y = "Cases", title = "Scatter Plot of Illness by Year and Pathogen Type")

## Save Graph
ggsave("time_v_illness_pathogen_type.png", path = "results")

# Pathogen Type + Etiology vs Illnesses Caused
ggplot(analysis_data_final %>% group_by(Pathogen_Type, Simplified_Etiology, Illnesses) %>% count() %>% mutate(true_illness_count = n * Illnesses), aes(x = Pathogen_Type, y = Illnesses, group = Simplified_Etiology, fill = Simplified_Etiology)) + geom_col() + labs(x = "Pathogen Type", y = "Cases", title = "Stacked Bar Chart of Illnesses by Pathogen Type and Etiology")

## Save Graph
ggsave("pathogen_type_etiology_v_illness.png", path = "results")

## Closer Look at Bacterial Etiologies
ggplot(analysis_data_final %>% group_by(Pathogen_Type, Simplified_Etiology, Illnesses) %>% filter(Pathogen_Type == "Bacteria"), aes(x = Simplified_Etiology)) + geom_bar() + labs(x = "Bacteria", y = "Cases", title = "Bar Chart of Bacteria Frequency")

ggsave("bacterial_etiology_v_outbreak_frequency.png", path = "results")

# States vs Illnesses

state_v_illness_analysis <- analysis_data_final %>% group_by(State) %>% mutate(all_illnesses = sum(Illnesses)) %>% count(all_illnesses) %>% select(State, all_illnesses) %>% arrange(desc(all_illnesses))

saveRDS(state_v_illness_analysis, file = "results/state_v_illness_analysis.rds")

rmarkdown::paged_table(state_v_illness_analysis)

# Food Type and Hospitalizations
ggplot(analysis_data_final %>% filter(`IFSAC Category` != "Multiple") %>% filter(`IFSAC Category` != "Other"), aes(x = `IFSAC Category`, y = Illnesses)) + geom_col() + labs(x = "Food Type", y = "Illnesses", title = "Illnesses by Food Type, Excluding 'Multiple' and 'Other'")

ggsave("food_type_v_illness_frequency.png", path = "results")

```
???

### 2023-02-24 code/processing_code/processingfile_v1.qmd
part 2 additional work

## Training Data
```{r}
training_data_fin <- training_data %>%
  select(!c(Ingredient, `Serotype/Genotype`)) %>%
  filter(grepl("Confirmed", Status))
```

@@

# checking the levels under each variable

:D

```{r}
skim(d1$Food)
#convert Food to factor
d1$Food=as.factor(d1$Food)
str(d1$Food)
#Since there are over 3000 different levels, we need to combine some food into the same category
d2=
```

checking the levels under `Species`
```{r}

# Checking the levels under `Location`
```{r}

## Data Manipulation
### Separate Rows to Address Coinfections
```{r}
analysis_data_fin <- analysis_data_fin %>%
  separate_rows(Etiology, sep = "; ")
```

@@

### Column for Type of Pathogen
```{r}
Pathogen_Bacteria <- c("Bacillus", "Brucella", "Campylobacter", "Clostridium", "Staphylococcus", "Escherichia", "Listeria", "Salmonella", "Vibrio", "Cholera", "Enterococcus", "Bacterium", "Shigella", "Yersinia")
Pathogen_Parasite <- c("Cryptosporidium", "Cyclospora", "Entamoeba", "Giardia", "Parasite", "Toxoplasma", "Trichinella")
Pathogen_Virus <- c("Astrovirus", "Norovirus", "Rotavirus")
Pathogen_Poison <- c("Poison", "Toxin", "Metal")

@@

## Check Data Structure Again
```{r}
skim(analysis_data_fin)
```

```{r}
rawdata1%>%filter(grepl("Confirmed", Status))
food=rawdata1%>%
  filter(!is.na(Food))%>%
  group_by(Food)%>%
  top_n(n=10)
food
species=rawdata1%>%
  filter(!is.na(Species))%>%
  mutate(Species=fct_collapse(Species,Norovirus=c("Norovirus","Norovirus genogroup I","Norovirus genogroup II","Norovirus unknown")))%>%
  mutate(Species=fct_lump(Species,11,other_level = NA))%>%
  count(Species)%>%
  arrange(desc(n))%>%
  filter(!is.na(Species))
species
```

All done, data is clean now.

Let's assign at the end to some final variable, this makes it easier to add further cleaning steps above.

```{r}
processeddata <- d4
```

# Save data

Finally, we save the clean data as RDS file.

```{r}
save_data_location <- here::here("data","processed_data","processeddata1.rds")
saveRDS(processeddata1, file = save_data_location)
```

# Notes

Removing anyone who had "faulty" or missing data is one approach. It's often not the best. based on your question and your analysis approach, you might want to do cleaning differently (e.g. keep individuals with some missing information).

>>>>

## Training Data
```{r}
training_data_fin <- training_data %>%
  select(!c(Ingredient, `Serotype/Genotype`)) %>%
  filter(grepl("Confirmed", Status))
```
# Data Manipulation
## Training Data
### Cleaning the "Species" Column
```{r}
unique(training_data_fin$Species)
# Separate the rows for coninfections
training_data_fin=training_data_fin%>%
  separate_rows(Species,sep=";")
# Create lists for pathogen types
Pathogen_Bacteria <- c("Bacillus", "Brucella", "Campylobacter", "Clostridium", "Staphylococcus", "Escherichia", "Listeria", "Salmonella", "Vibrio", "Cholera", "Enterococcus", "Bacterium", "Shigella", "Yersinia","Streptococcus")
Pathogen_Parasite <- c("Cryptosporidium", "Cyclospora", "Entamoeba", "Giardia", "Parasite", "Toxoplasma", "Trichinella")
Pathogen_Virus <- c("Astrovirus", "Norovirus", "Rotavirus","Hepatitis","Virus")
Pathogen_Poison <- c("Poison", "Toxin", "Metal","Pesticide","Ciguatoxin","Scombroidtoxin")
# simplify "Escherichia coli, Shiga toxin-producing" to "Escherichia coli"
# Create two new columns: Pathogen_type & Simplified_Etiology
training_data_fin=training_data_fin%>%
  mutate(Pathogen_Type = case_when(
    grepl(paste(Pathogen_Virus, 
                collapse = "|"), Species,ignore.case = TRUE) ~ "Virus",
    grepl(paste(Pathogen_Bacteria, 
                collapse = "|"), Species,ignore.case = TRUE) ~ "Bacteria",
    grepl(paste(Pathogen_Parasite, 
                collapse = "|"), Species,ignore.case = TRUE) ~ "Parasite",
    grepl(paste(Pathogen_Poison, 
                collapse = "|"), Species,ignore.case = TRUE) ~ "Poison"
    )) %>%
  mutate(Simplified_Etiology = case_when(
    Pathogen_Type == "Bacteria" ~ str_extract(Species, 
      "Bacillus|Brucella|Campylobacter|Clostridium|Staphylococcus|Escherichia|Listeria|Salmonella|Vibrio|Cholera|Enterococcus|Bacterium|Shigella|Yersinia|Streptococcus"),
    Pathogen_Type == "Parasite" ~ str_extract(Species, "Cryptosporidium|Cyclospora|Entamoeba|Giardia|Parasite|Toxoplasma|Trichinella"),
    Pathogen_Type == "Virus" ~ str_extract(Species, "Astrovirus|Norovirus|Rotavirus|Hepatitis"),
    Pathogen_Type == "Poison" ~ str_extract(Species, "Poison|Toxin|Heavy|pesticide|Ciguatoxin|Scombroid")
    )) %>%
      mutate(Simplified_Etiology = case_when(
        Simplified_Etiology == "Escherichia" ~ "E. coli",
        Simplified_Etiology == "Heavy" ~ "Heavy Metals",
        TRUE ~ Simplified_Etiology))
```

### Column for Location
```{r}
# Checking Location Names and Separate Multiple Names
unique(training_data_fin$Location)
training_data_fin=training_data_fin %>%
 separate_rows(Location,sep="; ")

The variable for location appears to be clean now.

### Make IFSAC Category Column for Training Data

Everything else not included in vectors will be categorized as "Other".

```{r}
# unique(analysis_data_fin$`IFSAC Category`)
# unique(training_data_fin$Food)
Desserts=c("Cookie","Cream","Canoli","Cake","Pudding")
Beef=c("Beef","Steak")
Poultry=c("Chicken","Turkey")
Pork=c("Pork","Sausage","Hot Dog","Deli","Spam","Strawberries","Blueberries","Pig","Ribs")
Dairy=c("Milk","Cheese","Parsley")
Crusteaceans=c("Lobster","Crab","Shrimp")
Fish=c("Fish","Tuna","Salmon")
Mollusks=c("Clams","Oysters")
Fruits=c("Tomato","Banana","Melon","Grapes","Mango")
`Grains-Beans`=c("Rice","Peas","Beans","wheat")
`Vegetable-Row-Crops`=c("Lettuce","artichokes","broccoli","asparagus","celeries","spinach")
`Root-underground-vegetables`=c("Roots","carrots","Beets","Onion","Tubers","yams","Garlic","Potatos","taro")
Sprouts=c("alfalfa","mung")
Salad=c("Salad")
# Other=c("Ice","Alcohol","Eggnog")
Game=c("Bear")
Herbs=c("basil","cilantro")
Fungi=c("Mushroom","portabellas")
`Nuts-Seeds`=c("peanuts","almonds","Seasames")
Eggs=c("Egg")
`Seeded-Vegetables`=c("Peppers","Squashes","Cucumbers","Okras","Corn")
`other meat`=c("Lamb")

training_data_fin <- training_data_fin %>%
  mutate(`IFSAC Category` = case_when(
    grepl("Salad",Food, ignore.case = TRUE) ~ "Salad",
    grepl(paste(Desserts, 
                collapse = "|"), Food,ignore.case = TRUE) ~ "Dessert",
    grepl(paste(Poultry,
                collapse = "|"), Food, ignore.case = TRUE) ~ "Poultry",
    grepl(paste(Beef,
                collapse = "|"), Food, ignore.case = TRUE) ~ "Beef",
    grepl(paste(Pork,
                collapse = "|"), Food, ignore.case = TRUE) ~ "Pork",
    grepl(paste(Dairy,
                collapse = "|"), Food, ignore.case = TRUE) ~ "Dairy",
    grepl(paste(Crusteaceans,
                collapse = "|"), Food, ignore.case = TRUE) ~ "Crusteaceans",
    grepl(paste(Fish,
                collapse = "|"), Food, ignore.case = TRUE) ~ "Fish",
    grepl(paste(Mollusks,
                collapse = "|"), Food, ignore.case = TRUE) ~ "Mollusks",
    grepl(paste(Fruits,
                collapse = "|"), Food, ignore.case = TRUE) ~ "Fruits",
    grepl(paste(`Grains-Beans`,
                collapse = "|"), Food, ignore.case = TRUE) ~ "Grains-Beans",
    grepl(paste(`Vegetable-Row-Crops`,
                collapse = "|"), Food, ignore.case = TRUE) ~ "Vegetable-Row-Crops",
    grepl(paste(`Root-underground-vegetables`,
                collapse = "|"), Food, ignore.case = TRUE) ~ "Root-underground-vegetables",
    grepl(paste(Sprouts,
                collapse = "|"), Food, ignore.case = TRUE) ~ "Sprouts",

    grepl(paste(Herbs,
                collapse = "|"), Food, ignore.case = TRUE) ~ "Herbs",
    grepl(paste(Fungi,
                collapse = "|"), Food, ignore.case = TRUE) ~ "Fungi",
    grepl(paste(`Nuts-Seeds`,
                collapse = "|"), Food, ignore.case = TRUE) ~ "Nuts-Seeds",
    grepl(paste(`Seeded-Vegetables`,
                collapse = "|"), Food, ignore.case = TRUE) ~ "Seeded-Vegetables",
    grepl("eggs",Food,ignore.case = TRUE)~"Eggs",
    grepl("Lamb",Food)~"Other Meat",
    TRUE ~ if_else(is.na(Food),"","Other")
   ))%>%
  mutate(`IFSAC Category` = na_if(`IFSAC Category`, ""))%>%
  relocate(`IFSAC Category`, .after = Food)
training_data_fin=training_data_fin%>%
  mutate(`IFSAC Category`=case_when(
    grepl(";",Food)~"Multiple",TRUE~`IFSAC Category`))
```

@@

### Separate Rows to Address Coinfections
```{r}
analysis_data_fin <- analysis_data_fin %>%
  separate_rows(Etiology, sep = "; ")
```

@@

### Column for Type of Pathogen
```{r}

@@

# Drop Rows Containing NA
```{r}
training_data_final=training_data_fin%>%
  drop_na()
analysis_data_final=analysis_data_fin%>%
  drop_na()

## Check Data Structure Again
```{r}
skim(training_data_final)
skim(analysis_data_final)
```

# Save Data

The cleaned datasets will be saved as RDS files.

```{r}
# Label Final Datasets as Cleaned/Processed
processed_training_data <- training_data_final
processed_analysis_data <- analysis_data_final

# Save Data
save_data_location1 <- here::here("data","processed_data","processed_training_data.rds")
saveRDS(processed_training_data, file = save_data_location1)

save_data_location2=here::here("data","processed_data","processed_analysis_data.rds")
saveRDS(processed_analysis_data, file = save_data_location2)
